---
title: "Spatial Cascade Monte Carlo Analysis"
subtitle: "Method, use and strategic results"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
config <- jsonlite::fromJSON("config.json")
results <- readRDS("assets/spatial_cascade_results.rds")
summary <- read.csv("assets/spatial_cascade_summary.csv")
type_summary <- read.csv("assets/spatial_cascade_production_type_summary.csv")

population <- results$population_size
population_source <- results$population_size_source
runs <- config$simulation_controls$monte_carlo_iterations
days <- config$model_parameters$simulation_days
allocation <- 100 * config$housing_attributes$barn_allocation_rate
protection <- 100 * config$housing_attributes$barn_protection_efficiency
type_mix <- paste(
  paste0(config$production_metrics$production_types$type, " ",
         round(100 * config$production_metrics$production_types$allocation_rate), "%"),
  collapse = ", "
)
total_average_capacity <- population *
  config$production_metrics$average_bird_yield
mean_loss_share <- 100 * summary$mean_production_loss / total_average_capacity
```

## Purpose

This report uses poultry farms as the practical example. It asks: if a harmful
event starts at a few farms, how widely might it spread, how much production
might be lost, and how much difference could sheltered housing make?

The model is useful for comparing protection strategies and understanding the
range of plausible outcomes. It is not a forecast for a specific real event.

## What a node means

A **node is one farm site**. In a real dataset, one row would represent one
farm. Each farm has:

- one latitude and longitude;
- one production type, such as eggs or meat;
- a sheltered or unsheltered housing status; and
- an amount of production that could be lost.

The current example uses `r population` farms. The population size comes from
**`r population_source`**. Average production is represented
by a value of `r format(config$production_metrics$average_bird_yield, big.mark = ",")`
per farm. This value could mean birds, tonnes, trays of eggs, revenue or another
useful capacity measure, as long as the same unit is used for every farm.

## What happens in one simulation run

One run is one possible outbreak scenario—not one farm and not one day.

For one run, the model:

1. Creates `r population` farm locations inside the configured area.
2. Assigns production type (`r type_mix`) and sheltered status to each farm.
3. Starts the event at three randomly selected farms.
4. Simulates spread and farm-state changes for `r days` days.
5. Records the farms that remain unaffected and the production lost.

Farms closer to an active farm face higher risk. Risk falls smoothly with
distance. Sheltered housing reduces that risk by the configured protection
effect. Production type does not change risk; it is used only to split the final
results.

## What the four farm states mean

Every farm moves through four stages:

1. **`r config$scenario_metadata$phase_a_label`** — the farm has not been
   affected.
2. **`r config$scenario_metadata$phase_b_label`** — the farm has been affected,
   but cannot yet affect other farms.
3. **`r config$scenario_metadata$phase_c_label`** — the farm can now affect
   nearby farms.
4. **`r config$scenario_metadata$phase_d_label`** — the event has run its active
   course and that farm's production is counted as lost.

This is a one-way path. A farm does not recover or return to production within
the current simulation period.

## What 500 Monte Carlo runs mean

The complete farm scenario is repeated **`r runs` times**. Each run creates new
farm locations, new sheltered assignments, new production types, new starting
farms and new random transmission outcomes.

The purpose is to see how much results vary across many plausible versions of
the same scenario. Each value in the loss histogram is the total loss from one
complete `r days`-day run.

More runs make averages, percentiles and charts more stable by reducing random
simulation noise. They do **not** make unrealistic assumptions accurate, and
500 runs do not mean there is a 500-day event. The results remain conditional on
the distances, timings and protection values set in `config.json`.

## Results

![](assets/spatial_cascade_loss_distribution.svg){fig-alt="Distribution of production capacity lost across Monte Carlo runs"}

Average lost production capacity was **`r format(round(summary$mean_production_loss), big.mark = ",")`**, or about
**`r round(mean_loss_share, 1)`%** of the population's expected starting
capacity. The middle 90% of simulated outcomes ranged from
**`r format(round(summary$loss_p05), big.mark = ",")`** to
**`r format(round(summary$loss_p95), big.mark = ",")`**.

The spread of the histogram is strategically important: one average number does
not capture the risk. Planning should allow for outcomes toward the upper end,
not only the centre of the distribution.

Sheltered farms finished unaffected in
**`r round(summary$sheltered_survival_pct, 1)`%** of cases on average, compared
with **`r round(summary$unsheltered_survival_pct, 1)`%** for unsheltered farms.
The protection difference was therefore
**`r round(summary$protection_benefit_percentage_points, 1)` percentage
points**.

Under the current assumptions, sheltering materially reduces exposure. The
strategic question is whether that improvement justifies the cost of increasing
the current `r round(allocation)`% allocation rate. That decision should be
tested with realistic protection costs and real locations before operational
use.

## Results by production type

![](assets/spatial_cascade_loss_by_production_type.svg){fig-alt="Production loss distributions split by production type"}

```{r}
#| label: production-type-table
#| tbl-cap: "Average results by production type"
display_type <- type_summary[, c(
  "production_type", "configured_allocation_rate",
  "average_farms_in_population", "average_farms_lost",
  "average_farms_lost_pct", "mean_total_production_loss",
  "sd_total_production_loss", "total_loss_p25", "total_loss_p75",
  "sheltered_survival_pct", "unsheltered_survival_pct"
)]
display_type$configured_allocation_rate <-
  paste0(round(100 * display_type$configured_allocation_rate), "%")
display_type$average_farms_in_population <-
  round(display_type$average_farms_in_population, 1)
display_type$average_farms_lost <- paste0(
  round(display_type$average_farms_lost, 1), " (",
  round(display_type$average_farms_lost_pct, 1), "%)"
)
display_type$average_farms_lost_pct <- NULL
display_type$mean_total_production_loss <-
  format(round(display_type$mean_total_production_loss), big.mark = ",")
display_type$sd_total_production_loss <-
  format(round(display_type$sd_total_production_loss), big.mark = ",")
display_type$total_loss_p25 <- paste0(
  format(round(display_type$total_loss_p25), big.mark = ","), "–",
  format(round(display_type$total_loss_p75), big.mark = ",")
)
display_type$total_loss_p75 <- NULL
display_type$sheltered_survival_pct <-
  round(display_type$sheltered_survival_pct, 1)
display_type$unsheltered_survival_pct <-
  round(display_type$unsheltered_survival_pct, 1)
names(display_type) <- c(
  "Production type", "Configured share", "Average farms represented",
  "Average farms lost (%)", "Mean total loss", "Loss SD",
  "25th to 75th percentile loss",
  "Sheltered survival (%)", "Unsheltered survival (%)"
)
knitr::kable(display_type, row.names = FALSE)
```

The loss values describe the **whole production-type population in one run**.
For example, the eggs result is the total loss across all egg farms in that
scenario, averaged across the `r runs` runs. It is not the loss of one average
egg farm.

The standard deviation shows how much total loss varies around the mean. The
25th to 75th percentile band contains the middle half of outcomes and is usually
easier to use for routine planning than one average. The existing 5th to 95th
percentiles in the CSV give a wider stress-testing range.

For a population view, focus on three measures together:

- **Average farms represented** shows the size of each category.
- **Average farms lost (%)** shows how many farm sites reach the final lost
  state, both as a count and as a share of that category.
- **Mean total loss and its band** show the production capacity exposed across
  the whole category.

The larger production category will normally show more total loss because it
contains more farms. Differences in survival between categories are sampling
variation, not a different transmission rule. The main strategic comparison
remains sheltered versus unsheltered within each category.

![](assets/spatial_cascade_survival_comparison.svg){fig-alt="Sheltered and unsheltered survival split by production type"}

## How to use it

1. Edit `config.json`.
2. Set the number of farms and simulation days. To use the number of rows in a
   farm CSV instead, set `use_row_count_for_population_size` to `true` and set
   `input_data.file` to the CSV path. Only the row count is used.
3. Set the map area and how quickly risk falls with distance.
4. Set the incubation and active periods.
5. Set average production per farm, production types and their shares.
6. Set the share of farms with sheltered housing and its expected protection.
   Production-type shares must sum to 1.
7. Run:

```{r}
#| eval: false
source("spatial_cascade_monte_carlo.R")
```

8. Review the SVG charts and summary files in `assets/`.
9. Render this report after the simulation finishes:

```bash
quarto render spatial_cascade_report.qmd
```

For quick testing, reduce `monte_carlo_iterations`. Increase it when the results
change noticeably between repeated executions. Several hundred runs are a
reasonable starting point for strategic comparisons of a model this size.

## What matters most

- **Distance decay** controls whether the event stays near the starting farms or
  spreads more widely.
- **Phase timing** controls how quickly farms become active and how long they can
  affect other farms.
- **Protection efficiency and allocation** determine how much of the population
  receives a reduced risk.
- **Farm locations** create uncertainty, which is why the model reports a range
  rather than one answer.

## Limits

The current locations are random within a rectangular map area, not actual farm
locations. The model does not prevent farms from being placed in water or other
unsuitable land. Transmission uses direct geographic distance and does not
include roads, terrain, prevailing wind, animal movements, cleaning practices,
farm contacts or response actions. Replace these assumptions and validate the
parameters before treating the output as a real farm-risk estimate.
