---
title: "Spatial Cascade Monte Carlo Analysis"
subtitle: "Method, use and strategic results"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
config <- jsonlite::fromJSON("config.json")
results <- readRDS("assets/spatial_cascade_results.rds")
summary <- read.csv("assets/spatial_cascade_summary.csv")
type_summary <- read.csv("assets/spatial_cascade_production_type_summary.csv")

population <- config$model_parameters$population_size
runs <- config$simulation_controls$monte_carlo_iterations
days <- config$model_parameters$simulation_days
allocation <- 100 * config$housing_attributes$barn_allocation_rate
protection <- 100 * config$housing_attributes$barn_protection_efficiency
type_mix <- paste(
  paste0(config$production_metrics$production_types$type, " ",
         round(100 * config$production_metrics$production_types$allocation_rate), "%"),
  collapse = ", "
)
total_average_capacity <- population *
  config$production_metrics$average_bird_yield
mean_loss_share <- 100 * summary$mean_production_loss / total_average_capacity
```

## Purpose

This model tests how a problem can spread through a spatial network and reduce
production capacity. The current configuration uses poultry and barn language,
but the same engine can represent farms, utilities, financial networks or urban
assets by changing `config.json`.

The model is useful for comparing protection strategies and understanding the
range of plausible outcomes. It is not a forecast for a specific real event.

## How the model works

Each simulation creates `r population` nodes at new random locations. Three
nodes begin in the active state. The model then runs for `r days` days.

Each node is also assigned a production type using the configured mix
(`r type_mix`). Production type is used only to split the reported results. It
does not change distance, transmission, protection or phase timing.

Every node moves through four stages:

1. **`r config$scenario_metadata$phase_a_label`** — not affected.
2. **`r config$scenario_metadata$phase_b_label`** — affected but not yet active.
3. **`r config$scenario_metadata$phase_c_label`** — able to affect nearby nodes.
4. **`r config$scenario_metadata$phase_d_label`** — its production value is
   counted as lost.

Risk falls as geographic distance increases. Sheltered nodes receive the
configured protection reduction. The simulation is repeated `r runs` times,
with a new geographic layout each time, so the results show a range rather than
one fixed answer.

## Results

![](assets/spatial_cascade_loss_distribution.svg){fig-alt="Distribution of production capacity lost across Monte Carlo runs"}

Average lost production capacity was **`r format(round(summary$mean_production_loss), big.mark = ",")`**, or about
**`r round(mean_loss_share, 1)`%** of the population's expected starting
capacity. The middle 90% of simulated outcomes ranged from
**`r format(round(summary$loss_p05), big.mark = ",")`** to
**`r format(round(summary$loss_p95), big.mark = ",")`**.

The spread of the histogram is strategically important: one average number does
not capture the risk. Planning should allow for outcomes toward the upper end,
not only the centre of the distribution.

Sheltered nodes finished unaffected in
**`r round(summary$sheltered_survival_pct, 1)`%** of cases on average, compared
with **`r round(summary$unsheltered_survival_pct, 1)`%** for unsheltered nodes.
The protection difference was therefore
**`r round(summary$protection_benefit_percentage_points, 1)` percentage
points**.

Under the current assumptions, sheltering materially reduces exposure. The
strategic question is whether that improvement justifies the cost of increasing
the current `r round(allocation)`% allocation rate. That decision should be
tested with realistic protection costs and real locations before operational
use.

## Results by production type

![](assets/spatial_cascade_loss_by_production_type.svg){fig-alt="Production loss distributions split by production type"}

```{r}
#| label: production-type-table
#| tbl-cap: "Average results by production type"
display_type <- type_summary[, c(
  "production_type", "configured_allocation_rate", "mean_production_loss",
  "sheltered_survival_pct", "unsheltered_survival_pct"
)]
display_type$configured_allocation_rate <-
  paste0(round(100 * display_type$configured_allocation_rate), "%")
display_type$mean_production_loss <-
  format(round(display_type$mean_production_loss), big.mark = ",")
display_type$sheltered_survival_pct <-
  round(display_type$sheltered_survival_pct, 1)
display_type$unsheltered_survival_pct <-
  round(display_type$unsheltered_survival_pct, 1)
names(display_type) <- c(
  "Production type", "Configured share", "Average loss",
  "Sheltered survival (%)", "Unsheltered survival (%)"
)
knitr::kable(display_type, row.names = FALSE)
```

The larger production category will normally show more total loss because it
contains more nodes. Differences in survival between categories are sampling
variation, not a different transmission rule. The main strategic comparison
remains sheltered versus unsheltered within each category.

![](assets/spatial_cascade_survival_comparison.svg){fig-alt="Sheltered and unsheltered survival split by production type"}

## How to use it

1. Edit `config.json`.
2. Set the scenario labels, population, duration, spatial decay, phase timing,
   node value, production-type names and shares, allocation rate, protection
   efficiency and map bounds. Production-type shares must sum to 1.
3. Run:

```{r}
#| eval: false
source("spatial_cascade_monte_carlo.R")
```

4. Review the SVG charts and summary files in `assets/`.
5. Render this report after the simulation finishes:

```bash
quarto render spatial_cascade_report.qmd
```

For quick testing, reduce `monte_carlo_iterations`. Use several hundred or more
runs for a stable strategic comparison.

## What matters most

- **Distance decay** controls how local or widespread the cascade becomes.
- **Phase timing** controls how quickly it grows and how long nodes remain
  active.
- **Protection efficiency and allocation** determine how much of the population
  receives a reduced risk.
- **Geographic layout** creates uncertainty, which is why the model reports a
  distribution rather than one answer.

## Limits

The current locations are random within a bounding box, not actual farm or
network locations. Transmission uses direct geographic distance and does not
include roads, terrain, prevailing wind, operating dependencies or response
actions. Replace these assumptions before treating the output as a real-world
risk estimate.
