---
title: "How Production Loss Could Spread Across New Zealand Farms"
subtitle: "A practical account of the scenario and its results"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
config <- jsonlite::fromJSON("config.json")
results <- readRDS("assets/spatial_cascade_results.rds")
summary <- read.csv("assets/spatial_cascade_summary.csv")
types <- read.csv("assets/spatial_cascade_production_type_summary.csv")
farms <- results$farms
runs <- results$monte_carlo_runs
days <- config$model_parameters$simulation_days
capacity_unit <- config$production_metrics$production_capacity_unit
if (is.null(capacity_unit) || !length(capacity_unit)) capacity_unit <- "birds"
mean_loss_share <- 100 * summary$mean_production_loss /
  sum(farms$production_yield)
protection_difference <- summary$protection_benefit_percentage_points
cull <- results$cull_comparison_summary
farm_risk <- results$farm_risk_summary
cull_farm_risk <- results$cull_farm_risk_summary
north_name <- grep("^North", unique(farms$barrier_group), value = TRUE)[1]
south_name <- grep("^South", unique(farms$barrier_group), value = TRUE)[1]
initial_starts <- config$simulation_controls$initial_phase_c_entities

island_story <- function(group_name) {
  rows <- farms$barrier_group == group_name
  group_size <- sum(rows)
  start_probability <- 100 * (1 -
    choose(nrow(farms) - group_size, initial_starts) /
      choose(nrow(farms), initial_starts))
  group_risk <- mean(farm_risk$probability_lost_pct[
    farm_risk$barrier_group == group_name])
  sheltered_share <- 100 * mean(farms$is_sheltered[rows])
  data.frame(group = group_name, farms = group_size,
    start_probability = start_probability, mean_lost_probability = group_risk,
    sheltered_share = sheltered_share)
}

island_comparison <- rbind(island_story(north_name), island_story(south_name))
north <- island_comparison[island_comparison$group == north_name, ]
south <- island_comparison[island_comparison$group == south_name, ]
```

## The story this model tells

This analysis asks a practical question: if a harmful event begins on a small
number of farms, how far could it spread and how much production could be lost?
It follows the real farm locations and attributes supplied in
`r config$input_data$file`.

The current population contains **`r format(nrow(farms), big.mark = ",")` farms**.
Each dot on the map is one of those farms. Its location, region,
production type, shelter category and production capacity stay the same in
every scenario. Production type is used to describe the results. Shelter also
changes the modelled chance of a farm being reached.

At the start of a scenario, every farm is unaffected. The event begins at
`r config$simulation_controls$initial_phase_c_entities` selected farms and the
model follows the next `r days` days. A farm can be reached, spend time in an
incubation period, become able to affect other farms, and finally have its
production counted as lost. Nearby farms are more likely to be reached than
distant farms, although the outcome still includes chance.

## Why the scenario is repeated

One run gives only one possible chain of events. This analysis repeats the
whole scenario **`r format(runs, big.mark = ",")` times** over the same farm
population. The starting locations and the chance outcomes can differ from one
run to the next. Looking across all runs shows which outcomes are common and
which sit at the more severe end of the range.

More runs make the pattern steadier, but they do not make the assumptions more
realistic. The value of the results still depends on the farm data and the
settings used.

## The production base used in every scenario

The simulation starts from the same national production base in every run.
The table provides context before any loss is modelled. The farm-level figures
describe how capacity is distributed across farms; they are not simulation
results.

```{r}
#| label: production-reference-table
#| tbl-cap: "Production-capacity reference for the fixed farm population"
capacity_reference_row <- function(label, rows) {
  capacity <- farms$production_yield[rows]
  data.frame(
    `Production type` = label,
    Farms = length(capacity),
    `Total capacity` = sum(capacity),
    `Share of national capacity` = 100 * sum(capacity) /
      sum(farms$production_yield),
    `Average per farm` = mean(capacity),
    `Median per farm` = median(capacity),
    `Middle half per farm` = paste0(
      format(round(unname(quantile(capacity, .25))), big.mark = ","), " to ",
      format(round(unname(quantile(capacity, .75))), big.mark = ",")),
    `Minimum to maximum per farm` = paste0(
      format(round(min(capacity)), big.mark = ","), " to ",
      format(round(max(capacity)), big.mark = ",")),
    check.names = FALSE)
}
capacity_reference <- do.call(rbind, c(
  lapply(sort(unique(farms$production_type)), function(type) {
    capacity_reference_row(type, farms$production_type == type)
  }),
  list(capacity_reference_row("Grand total", rep(TRUE, nrow(farms))))))
capacity_reference$`Total capacity` <- paste(
  format(round(capacity_reference$`Total capacity`), big.mark = ","), capacity_unit)
capacity_reference$`Share of national capacity` <- paste0(
  round(capacity_reference$`Share of national capacity`, 1), "%")
capacity_reference$`Average per farm` <- paste(
  format(round(capacity_reference$`Average per farm`), big.mark = ","), capacity_unit)
capacity_reference$`Median per farm` <- paste(
  format(round(capacity_reference$`Median per farm`), big.mark = ","), capacity_unit)
capacity_reference$`Middle half per farm` <- paste(
  capacity_reference$`Middle half per farm`, capacity_unit)
capacity_reference$`Minimum to maximum per farm` <- paste(
  capacity_reference$`Minimum to maximum per farm`, capacity_unit)
knitr::kable(capacity_reference, row.names = FALSE)
```

The **Grand total** row gives the full capacity at risk. Its average, median and
ranges describe individual farms across the complete input population.

## What happened across the whole farm population

![Production lost and remaining across `r format(runs, big.mark = ",")` matched `r days`-day simulations of all `r format(nrow(farms), big.mark = ",")` farms, shown separately with no culling and with culling. Each thin vertical bar is one complete scenario and always totals 100%. Within each panel, scenarios are sorted from lower to higher loss; their left-to-right order is not time or run order.](assets/spatial_cascade_loss_distribution.svg){#fig-loss-distribution fig-alt="Faceted one hundred percent stacked bars comparing production lost and remaining in matched scenarios with no culling and with culling"}

Each bar represents one possible national outcome. The orange share is
production lost and the grey share is production remaining. A bar that is 80%
orange means that scenario ended with 80% of national capacity lost. The two
panels use the same percentage scale. Comparing their orange areas shows how
the culling response shifts the range of national outcomes. The sorting is
done separately in each panel, so bars in the same horizontal position are not
necessarily the same matched run.

Across the repeated scenarios, average total loss was
**`r format(round(summary$mean_production_loss), big.mark = ",")` `r capacity_unit`**, about
**`r round(mean_loss_share, 1)`%** of the fixed population's production
capacity. Nine out of ten scenarios finished between
**`r format(round(summary$loss_p05), big.mark = ",")` and
`r format(round(summary$loss_p95), big.mark = ",")` `r capacity_unit` lost**.

This is a widespread-loss result under the current settings. The average is
useful for a central planning case, while the upper end of the range shows what
a more severe outcome could look like. Because loss is already high in most
runs, moderate changes to protection may be difficult to distinguish unless
the spread and timing settings are well supported by evidence.

## What changes when culling is used

The analysis also repeats every scenario with a culling response. This is a
fair comparison: the no-control and culling versions begin at the same farms
and use the same prepared chance values. The only intended difference is the
response.

The current policy reaches **`r round(cull$coverage_pct)`% of active farms**,
then culls each covered farm **`r cull$response_delay_days` day** after it
becomes active. There is no separate daily limit: once the response delay has
passed, every identified farm is culled.

![Total production-capacity loss across matched scenarios with no control and with the configured culling response. Culling loss includes the production removed through culling as well as farms otherwise reaching the lost state.](assets/spatial_cascade_cull_comparison.svg){#fig-cull-comparison fig-alt="Comparison of total production capacity loss with no control and with culling"}

Without control, average total loss was
**`r format(round(cull$mean_baseline_total_loss), big.mark = ",")` `r capacity_unit`**.
With culling, average total loss was
**`r format(round(cull$mean_cull_total_loss), big.mark = ",")` `r capacity_unit`**.
That culling total already includes about
**`r format(round(cull$mean_direct_cull_loss), big.mark = ",")` `r capacity_unit`**
removed directly through culling.

After counting that direct loss, the response avoided an average of
**`r format(round(cull$mean_net_loss_avoided), big.mark = ",")` `r capacity_unit`**,
or **`r round(cull$net_loss_reduction_pct, 1)`% of the no-control loss**. Total
loss was lower with culling in **`r round(cull$runs_with_lower_total_loss_pct, 1)`%
of the matched scenarios**. The average number of affected farms fell from
`r round(cull$mean_baseline_farms_affected, 1)` to
`r round(cull$mean_cull_farms_affected, 1)`, while an average of
`r round(cull$mean_farms_culled, 1)` farms were culled.

This result shows the trade-off clearly. Culling accepts a smaller, immediate
production loss to prevent a much larger chain. It does not mean this exact
result would occur in practice. The benefit depends on finding active farms,
reaching `r round(cull$coverage_pct)`% coverage, acting within
`r cull$response_delay_days` day.
Those assumptions should be tested against realistic response capability.

## How the result differs by production type

![Total production capacity lost for each production type across the repeated scenarios. Each box shows the range of whole-population loss for that category, not the loss of an average farm.](assets/spatial_cascade_loss_by_production_type.svg){#fig-loss-by-type fig-alt="Distribution of total production capacity lost by production type"}

Production type does not change how the event spreads in this model. It is used
to show where the production consequences fall. A category with more farms or
more capacity will generally have a larger total loss.

```{r}
#| label: type-table
#| tbl-cap: "What the repeated scenarios mean for each production type"
make_type_rows <- function(scenario, loss_log, count_log) {
  rows <- lapply(colnames(loss_log), function(type) {
    in_type <- farms$production_type == type
    data.frame(Scenario = scenario, `Production type` = type, check.names = FALSE,
      Farms = sum(in_type), `Population share` = 100 * mean(in_type),
      `Average farms lost` = mean(count_log[, type]),
      `Average farms lost pct` = 100 * mean(count_log[, type]) / sum(in_type),
      `Average total capacity lost` = mean(loss_log[, type]),
      `Loss p25` = unname(quantile(loss_log[, type], .25)),
      `Loss p75` = unname(quantile(loss_log[, type], .75)))
  })
  total_loss <- rowSums(loss_log)
  total_count <- rowSums(count_log)
  rows[[length(rows) + 1L]] <- data.frame(
    Scenario = scenario, `Production type` = "Total", check.names = FALSE,
    Farms = nrow(farms), `Population share` = 100,
    `Average farms lost` = mean(total_count),
    `Average farms lost pct` = 100 * mean(total_count) / nrow(farms),
    `Average total capacity lost` = mean(total_loss),
    `Loss p25` = unname(quantile(total_loss, .25)),
    `Loss p75` = unname(quantile(total_loss, .75)))
  do.call(rbind, rows)
}

display <- rbind(
  make_type_rows("No culling", results$production_loss_by_type_log,
                 results$lost_count_by_type_log),
  make_type_rows("Culling", results$cull_production_loss_by_type_log,
                 results$cull_lost_count_by_type_log))
display$`Population share` <- paste0(round(display$`Population share`, 1), "%")
display$`Average farms lost` <- paste0(round(display$`Average farms lost`, 1),
  " (", round(display$`Average farms lost pct`, 1), "%)")
display$`Average farms lost pct` <- NULL
display$`Average total capacity lost` <- paste(
  format(round(display$`Average total capacity lost`), big.mark = ","), capacity_unit)
display$`Middle half of results` <- paste0(
  format(round(display$`Loss p25`), big.mark = ","), " to ",
  format(round(display$`Loss p75`), big.mark = ","), " ", capacity_unit)
display$`Loss p25` <- display$`Loss p75` <- NULL
knitr::kable(display, row.names = FALSE)
```

The “middle half” column leaves out the lowest quarter and highest quarter of
the runs. It gives a simple view of the more usual range without implying that
the more severe results cannot happen. The **Total** row describes the whole
farm population within each response scenario; it is not the sum of category
percentages.

## What the shelter comparison says

![Average share of sheltered and unsheltered farms that remain unaffected at the end of the scenario, shown by production type.](assets/spatial_cascade_survival_comparison.svg){#fig-shelter-comparison fig-alt="Comparison of sheltered and unsheltered farm survival by production type"}

Sheltered survival averaged **`r round(summary$sheltered_survival_pct, 1)`%**,
compared with **`r round(summary$unsheltered_survival_pct, 1)`%** for
unsheltered farms. The difference was
**`r round(summary$protection_benefit_percentage_points, 1)` percentage
points** under the configured assumptions.

```{r}
#| results: asis
if (protection_difference < 0) {
  cat(paste0(
    "In these results, sheltered farms have a slightly lower raw survival ",
    "rate. This does **not** mean shelter causes harm. The sheltered farms may ",
    "simply be in more exposed places or may have been chosen as starting ",
    "farms more often. The chart describes this farm population; it does not ",
    "prove what shelter caused.\n"
  ))
} else {
  cat(paste0(
    "Sheltered farms have a higher raw survival rate in these results. Some of ",
    "that difference may still come from where those farms are located, so it ",
    "should not be treated as proof of the shelter effect on its own.\n"
  ))
}
```

The cleanest future test would compare the same farms and the same starting
conditions twice: once with shelter protection and once without it. That would
separate the modelled shelter effect from differences in farm location.

## Where the exposure sits

![Farm exposure across New Zealand without control and with the configured culling response. Both panels use the same farms, bubble sizes and colour scale. Larger bubbles have more production capacity. Colour shows how often each farm was counted as lost across the repeated scenarios.](assets/spatial_cascade_farm_risk_map.svg){#fig-farm-risk fig-alt="Side-by-side maps of fixed New Zealand farm exposure without control and with culling"}

The left map shows the no-control result and the right map shows the matched
culling result. Both bring the two islands into one view. Each bubble is one
farm at its fixed location, and larger bubbles have more production capacity.
Because the colour scale is shared, the two panels can be compared directly.

In the no-control panel, colour shows how often a farm reached the lost state
through the normal modelled chain. In the culling panel, a farm is also counted
as lost when it is culled. This is important: a coloured farm on the right may
represent an early, deliberate loss that helped prevent many other farms from
being reached.

For example, a value of 40% means that the farm reached the lost state in 40
out of every 100 simulated scenarios under the current settings and starting
location method. It does not mean the farm is certain to lose 40% of its own
production, and it is not a forecast that an event will occur. It is a
scenario-based exposure measure: a high value identifies a farm that is often
reached when many possible cascades are tested.

For example, 40% means the farm was counted as lost in 40 out of every 100
modelled scenarios for that panel. It does not mean that the farm loses 40% of
its own production, and it is not a forecast that the event will occur. A high
value simply marks a farm that is frequently exposed under the chosen settings
and starting method.

Across all farms, the average lost-state probability falls from
**`r round(mean(farm_risk$probability_lost_pct), 1)`% without control** to
**`r round(mean(cull_farm_risk$probability_lost_pct), 1)`% with culling**.
On average, farms are directly culled in
**`r round(mean(cull_farm_risk$probability_culled_pct), 1)`% of scenarios**.
The right panel should therefore be read as a mix of farms removed by policy
and the smaller chain that continues despite that response.

The supporting farm-risk CSV also records how often each farm started a
scenario, how often it was affected, how often it was lost, and its expected
contribution to total loss. In random-start mode, some farms will be selected
as starting points more often than others just by chance. The current setup
selects each farm about
`r round(runs * config$simulation_controls$initial_phase_c_entities / nrow(farms), 1)`
times on average.

### Why the North Island looks more exposed

In the current results, the average chance of reaching the lost state is
**`r round(north$mean_lost_probability, 1)`% for North Island farms** and
**`r round(south$mean_lost_probability, 1)`% for South Island farms**. The main
reason is how the scenarios begin.

There are `r north$farms` North Island farms and `r south$farms` South Island
farms in the fixed population. With `r initial_starts` starting farms chosen at
random, at least one North Island farm starts the event in about
**`r round(north$start_probability, 1)`% of runs**. The South Island receives a
starting farm in about **`r round(south$start_probability, 1)`% of runs**.

The cross-island setting is currently 0, so an event cannot move from one
island to the other. If a run has no South Island starting farm, every South
Island farm stays unaffected in that run. The larger North Island population
therefore gives the North Island many more opportunities to be involved.

This should not be read as proof that North Island farms are naturally more
vulnerable. It is mainly a result of the number and location of farms, random
starting points, and the blocked island crossing. The shelter shares are also
fairly similar—`r round(north$sheltered_share, 1)`% in the North Island and
`r round(south$sheltered_share, 1)`% in the South Island—so shelter mix is not
the main explanation here. These figures will update automatically when the
full farm population is loaded and the analysis is rerun.

## What happens between the islands

The farm data identify which island each farm belongs to. Farms on the same
island have no extra barrier. For farms on different islands, the model applies
the cross-island setting after allowing for distance and shelter.

The current value is
**`r config$model_parameters$cross_barrier_transmission_multiplier`**. A value
of 0 blocks spread between islands. A value such as 0.05 leaves a small
cross-island pathway, while 1 removes the extra island penalty but still allows
distance to reduce spread. A value above 0 should only be used when there is a
credible route, such as ferry movements, shared equipment or supplier links.

The grey New Zealand shape and optional regional lines are there to make the
map easier to read. They do not change the model.

## What to take from these results

Under the current settings, the main message is that loss is widespread in
most scenarios. The range across runs matters because it shows that the final
loss is not fixed, even when the farm population stays the same. The map then
adds the local view by showing which farms are repeatedly exposed.

These results are best used to compare plausible settings and identify where
better evidence is needed. Straight-line distance cannot represent every real
pathway. Animal movements, staff, vehicles, shared equipment, wind, cleaning
practices and response actions may all change the real pattern. The model
should support discussion, not replace operational judgement.

## Running the analysis again

Start with the complete farm table in `data/farms.csv`. If its column names
change, update the mappings in `config.json`. Choose how starting farms are
selected, review the timing, distance, shelter, island and run settings, then
run `source("spatial_cascade_monte_carlo.R")`. The refreshed charts and tables
will be written to `assets/`. Finish by running
`quarto render spatial_cascade_report.qmd` to rebuild this report.

# Appendix: How timing and distance shape spread

![The left panel follows one farm after it is reached and shows when it can spread, when it would normally move to lost production, and when a covered farm would be culled. The right panel shows how one transmission opportunity becomes less likely as the distance between two farms increases.](assets/spatial_cascade_timing_distance_concept.svg){#fig-timing-distance fig-alt="Two-panel concept figure showing farm state timing and spatial transmission decay"}

The two parts of the model work together. **Timing controls when a farm can
spread. Distance controls how likely it is to reach another farm while it is
active.**

The left panel begins when a farm is reached. It spends
`r config$model_parameters$incubation_period_days` days incubating and cannot
spread during that period. It then becomes active for up to
`r config$model_parameters$active_period_days` days. Without control, it moves
to lost production at the end of that active period. If the farm is covered by
the culling policy, it is culled
`r config$cull_response$response_delay_days` day after becoming active and can
no longer spread. Coverage still matters: only
`r round(100 * config$cull_response$coverage)`% of active farms are selected
for that response under the current settings.

The right panel begins with one active farm and one unaffected farm. At short
distances, the model gives that transmission opportunity a higher chance. The
chance falls smoothly as distance increases. Shelter lowers the whole curve
for a sheltered farm. The different-island curve also includes the island
setting; with the current value of
`r config$model_parameters$cross_barrier_transmission_multiplier`, it stays at
zero because cross-island spread is blocked.

This curve describes one opportunity from one active source. When several
active farms can reach the same farm on the same day, the model combines those
opportunities, so the farm's overall chance of being reached can be higher than
one line alone suggests.
