---
title: "Fixed-Farm Spatial Cascade Analysis"
subtitle: "Method, use and strategic results"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
config <- jsonlite::fromJSON("config.json")
results <- readRDS("assets/spatial_cascade_results.rds")
summary <- read.csv("assets/spatial_cascade_summary.csv")
types <- read.csv("assets/spatial_cascade_production_type_summary.csv")
farms <- results$farms
runs <- results$monte_carlo_runs
days <- config$model_parameters$simulation_days
capacity_unit <- config$production_metrics$production_capacity_unit
if (is.null(capacity_unit) || !length(capacity_unit)) capacity_unit <- "birds"
mean_loss_share <- 100 * summary$mean_production_loss /
  sum(farms$production_yield)
protection_difference <- summary$protection_benefit_percentage_points
```

## Practical purpose

The model asks what could happen if a harmful event starts at one or more farms
and spreads through a known farm population. It estimates which farms are most
often affected, how much production could be lost, and whether sheltered farms
perform better than unsheltered farms.

The results describe the fixed farms in `r config$input_data$file`. They are no
longer based on a newly invented geography in every run.

## What one node means

One node is one farm row in the input table. The current table contains
**`r nrow(farms)` fixed farms**. Each farm keeps the same:

- farm ID and geographic location;
- island or other transmission-barrier group;
- region and production type;
- sheltered status; and
- production capacity.

These details do not change between Monte Carlo runs. Production type is used
only to report results. Shelter is different: it reduces transmission risk.

## What changes between runs

In each run, all farms reset to unaffected. The event then starts at farms
selected according to `starting_location_mode` in `config.json`.

- **random** selects `r config$simulation_controls$initial_phase_c_entities`
  starting farms for each of the configured runs;
- **each_farm** makes every farm the starting farm repeatedly, providing a fair
  comparison of which origin creates the largest cascade.

Transmission outcomes then vary randomly for `r days` days. Distances and farm
attributes remain fixed. The model applies the configured transmission penalty
when the source and target farms are in different barrier groups.

The four farm states are:

1. **Unaffected** — the farm has not been reached.
2. **Incubating** — reached, but not yet able to affect other farms.
3. **Active** — able to affect nearby farms.
4. **Lost production capacity** — production is counted as lost.

## What the number of runs means

This analysis completed **`r format(runs, big.mark = ",")` alternative outbreak
scenarios** over the same farms. A run is not one farm or one day. It is one
complete `r days`-day scenario.

More runs reduce random simulation noise and make percentages more stable. They
do not correct unrealistic transmission assumptions. In random-start mode,
check `times_selected_as_start` in the farm-risk CSV: some farms may be selected
more often than others. Use `each_farm` mode when origin-to-origin comparison is
the main question.

## Population results

![Total production-capacity loss across `r format(runs, big.mark = ",")` complete `r days`-day simulations of all `r format(nrow(farms), big.mark = ",")` farms in the fixed input population. The horizontal axis is the total `r capacity_unit` lost across the whole population in one run; the vertical axis is the number of runs with a similar loss. The dashed line is the average across runs.](assets/spatial_cascade_loss_distribution.svg){#fig-loss-distribution fig-alt="Histogram of total production capacity lost across complete Monte Carlo runs over the full fixed farm population"}

Read one bar as: **this many complete simulated scenarios ended with a total
population loss in this range**. The value is summed across all farms; it is
not the loss for an average farm. The distribution describes the range of
losses once a modelled event starts. It does not estimate the chance of a real
event starting.

Average total loss was
**`r format(round(summary$mean_production_loss), big.mark = ",")` `r capacity_unit`**, about
**`r round(mean_loss_share, 1)`%** of the fixed population's production
capacity. The 5th to 95th percentile range was
**`r format(round(summary$loss_p05), big.mark = ",")` to
`r format(round(summary$loss_p95), big.mark = ",")` `r capacity_unit`**.

The average represents about `r round(mean_loss_share, 1)`% of the fixed
population's capacity, so the current settings produce a widespread cascade.
This is strategically important: when most farms are reached, the model may
have limited ability to distinguish moderate protection options. The spatial
decay and timing assumptions should be calibrated before operational use.

The range matters more than the average alone. Routine planning can use the
middle of the distribution; resilience planning should also consider the upper
tail.

## Results by production type

![](assets/spatial_cascade_loss_by_production_type.svg){fig-alt="Total population loss distributions by production type"}

```{r}
#| label: type-table
#| tbl-cap: "Fixed-population results by production type"
display <- types[, c("production_type", "farms_in_population",
  "farms_in_population_pct", "average_farms_lost",
  "average_farms_lost_pct", "mean_total_production_loss",
  "sd_total_production_loss", "total_loss_p25", "total_loss_p75")]
display$farms_in_population_pct <- paste0(round(display$farms_in_population_pct, 1), "%")
display$average_farms_lost <- paste0(round(display$average_farms_lost, 1),
  " (", round(display$average_farms_lost_pct, 1), "%)")
display$average_farms_lost_pct <- NULL
display$mean_total_production_loss <- format(round(display$mean_total_production_loss), big.mark = ",")
display$sd_total_production_loss <- format(round(display$sd_total_production_loss), big.mark = ",")
display$total_loss_p25 <- paste0(format(round(display$total_loss_p25), big.mark = ","),
  " to ", format(round(display$total_loss_p75), big.mark = ","))
display$total_loss_p75 <- NULL
names(display) <- c("Production type", "Farms", "Population share",
  "Average farms lost (%)", "Mean total loss", "Loss SD",
  "25th to 75th percentile loss")
knitr::kable(display, row.names = FALSE)
```

These are population totals, not average-farm losses. A larger category will
usually show more total loss because it contains more farms. Production type
does not alter transmission.

![](assets/spatial_cascade_survival_comparison.svg){fig-alt="Sheltered and unsheltered survival by production type"}

Sheltered survival averaged **`r round(summary$sheltered_survival_pct, 1)`%**,
compared with **`r round(summary$unsheltered_survival_pct, 1)`%** for
unsheltered farms. The difference was
**`r round(summary$protection_benefit_percentage_points, 1)` percentage
points** under the configured assumptions.

```{r}
#| results: asis
if (protection_difference < 0) {
  cat(paste0(
    "Sheltered farms have lower raw survival in this run. This does **not** ",
    "mean shelter increases transmission. Shelter status and geography are ",
    "fixed, so sheltered farms may be concentrated in more exposed locations ",
    "or selected as starting farms more often. This is an observed population ",
    "comparison, not a causal estimate of shelter effectiveness.\n"
  ))
} else {
  cat(paste0(
    "Sheltered farms have higher raw survival in this fixed population. The ",
    "difference is still an observed comparison and may partly reflect where ",
    "sheltered farms are located.\n"
  ))
}
```

To isolate the effect of shelter, a future paired analysis should run the same
farms, starting locations and random transmission rolls twice: once with the
configured shelter protection and once with protection switched off. The
difference between those paired outcomes would be a cleaner protection-effect
estimate.

## Which farms are most exposed

![](assets/spatial_cascade_farm_risk_map.svg){fig-alt="Fixed farm locations coloured by probability of reaching the lost state"}

The map shows both North and South Island farms together, without separate
panels. The grey shape is New Zealand and, when enabled, the fine grey lines
show regional council boundaries. **Each point is one fixed farm** at its
actual longitude and latitude.
Point size represents that farm's production capacity. Point colour represents
the percentage of all Monte Carlo runs in which that farm reached **Lost
production capacity** by the end of the scenario.

For example, a value of 40% means that the farm reached the lost state in 40
out of every 100 simulated scenarios under the current settings and starting
location method. It does not mean the farm is certain to lose 40% of its own
production, and it is not a forecast that an event will occur. It is a
scenario-based exposure measure: a high value identifies a farm that is often
reached when many possible cascades are tested.

The map and `assets/spatial_cascade_farm_risk.csv` report for every farm:

- how often it was selected as a starting farm;
- probability of being affected;
- probability of reaching the final lost state; and
- expected contribution to total production loss.

In random-start mode these probabilities describe the configured mixture of
starting locations. They are not intrinsic farm scores independent of the start
sampling design. With `r runs` runs and
`r config$simulation_controls$initial_phase_c_entities` starts per run, a farm
is selected about
`r round(runs * config$simulation_controls$initial_phase_c_entities / nrow(farms), 1)`
times on average, but the actual count varies. Use `each_farm` mode for balanced
origin comparisons.

## Controlling transmission between islands

The model uses the `barrier_group` field in the fixed farm table. In the
current data this field is the island. Transmission within the same island has
no extra penalty. Transmission between different islands is multiplied by
`cross_barrier_transmission_multiplier` from `config.json`, after the normal
distance decay and shelter protection are applied.

The current value is **`r config$model_parameters$cross_barrier_transmission_multiplier`**,
so cross-island transmission is blocked. This means the model was already
keeping island cascades separate; the new setting simply makes that assumption
visible and controllable.

- `0` blocks transmission between islands.
- `0.05` retains 5% of the otherwise calculated transmission probability.
- `1` applies no extra island penalty, although geographic distance still
  reduces the probability.

Use a small value above zero only when there is a credible cross-island route,
such as ferry movements, shared equipment or supplier links. This single
multiplier is a planning simplification. For operational work, known movement
routes and biosecurity controls should be represented directly where data are
available.

## Map boundary option

The map uses the official Stats NZ 2025 regional council boundary file stored
in `data/nz_regional_council_2025.geojson`. In `config.json`, set
`map_options.show_region_boundaries` to `true` to show region outlines or
`false` to retain the New Zealand land shape without the internal region lines.
This setting changes only the presentation; it does not change transmission.

## How to use it

1. Replace or edit `data/farms.csv`.
2. Update `input_data.column_mappings` in `config.json` if column names differ.
3. Choose `random` or `each_farm` starting-location mode.
4. Set phase durations, spatial decay, the cross-barrier multiplier, protection
   efficiency, days and runs.
5. Run `source("spatial_cascade_monte_carlo.R")`.
6. Review the charts and CSV files in `assets/`.
7. Render this report with `quarto render spatial_cascade_report.qmd`.

## Strategic limits

Results are conditional on the farms and model settings supplied. Direct
geographic distance does not capture roads, wind, animal movements, shared
suppliers, cleaning practices or response actions. Farm-level probabilities
should not be used operationally until the fixed data and transmission settings
have been validated with subject-matter experts.
