---
title: "How Bird Loss Could Spread Across New Zealand Farms"
subtitle: "A practical comparison of no culling and culling"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{=html}
<style>
.table-note,
.figure-note {
  margin-top: -0.55rem;
  margin-bottom: 1.25rem;
  color: #7a8791;
  font-size: 0.82rem;
  line-height: 1.35;
}
.population-loss-table td[rowspan] {
  vertical-align: middle;
  font-weight: 600;
  background: #f5f7f8;
}
.key-findings {
  margin: 1rem 0 1.5rem;
  padding: 1rem 1.2rem 0.35rem;
  background: #fff7d6;
  border: 1px solid #eadb9a;
  border-left: 5px solid #d8bd55;
  border-radius: 0.35rem;
}
.key-findings-title {
  margin-bottom: 0.65rem;
  color: #5f5223;
  font-weight: 700;
}
</style>
```

```{r}
#| label: setup
config <- jsonlite::fromJSON("config.json")
results <- readRDS("assets/spatial_cascade_results.rds")
summary <- read.csv("assets/spatial_cascade_summary.csv")
farms <- results$farms
runs <- results$monte_carlo_runs
days <- config$model_parameters$simulation_days
capacity_unit <- config$production_metrics$production_capacity_unit
if (is.null(capacity_unit) || !length(capacity_unit)) capacity_unit <- "birds"
thousand_birds <- function(x, digits = 1) {
  format(round(x / 1000, digits), nsmall = digits, big.mark = ",", trim = TRUE)
}
mean_loss_share <- 100 * summary$mean_production_loss /
  sum(farms$production_yield)
cull <- results$cull_comparison_summary
farm_risk <- results$farm_risk_summary
cull_farm_risk <- results$cull_farm_risk_summary
cull_national_loss_pct <- 100 * cull$mean_cull_total_loss /
  sum(farms$production_yield)
cull_average_bird_survival <- results$cull_average_bird_survival
north_name <- grep("^North", unique(farms$barrier_group), value = TRUE)[1]
south_name <- grep("^South", unique(farms$barrier_group), value = TRUE)[1]
initial_starts <- config$simulation_controls$initial_phase_c_entities

island_story <- function(group_name) {
  rows <- farms$barrier_group == group_name
  group_size <- sum(rows)
  start_probability <- 100 * (1 -
    choose(nrow(farms) - group_size, initial_starts) /
      choose(nrow(farms), initial_starts))
  group_risk <- mean(farm_risk$probability_lost_pct[
    farm_risk$barrier_group == group_name])
  sheltered_share <- 100 * mean(farms$is_sheltered[rows])
  data.frame(group = group_name, farms = group_size,
    start_probability = start_probability, mean_lost_probability = group_risk,
    sheltered_share = sheltered_share)
}

island_comparison <- rbind(island_story(north_name), island_story(south_name))
north <- island_comparison[island_comparison$group == north_name, ]
south <- island_comparison[island_comparison$group == south_name, ]
cull_north_probability <- mean(cull_farm_risk$probability_lost_pct[
  cull_farm_risk$barrier_group == north_name])
cull_south_probability <- mean(cull_farm_risk$probability_lost_pct[
  cull_farm_risk$barrier_group == south_name])
production_types <- sort(unique(farms$production_type))
cull_type_loss_pct <- vapply(production_types, function(type) {
  type_birds <- sum(farms$production_yield[farms$production_type == type])
  expected_loss <- sum(cull_farm_risk$expected_loss_contribution[
    cull_farm_risk$production_type == type])
  100 * expected_loss / type_birds
}, numeric(1))
cull_type_loss_order <- order(cull_type_loss_pct, decreasing = TRUE)
cull_type_loss_comparison <- paste(
  paste0(production_types[cull_type_loss_order], ": ",
         round(cull_type_loss_pct[cull_type_loss_order], 1), "%"),
  collapse = "; ")
```

## What this analysis tells us

Each modelled outbreak begins at
`r config$simulation_controls$initial_phase_c_entities` farms selected at
random from the national farm population. The analysis asks how many birds
could be lost, where those losses could occur, and how much difference a
culling response could make.

The results tell a national story first, then show which production types and
places carry the loss. The same possible events are considered with no culling
and with a culling response, making the practical value of the response easier
to see. The simulation method and guidance on using the analysis are in the
appendix.

## The main findings point to early action

::: {.key-findings}

<div class="key-findings-title">Key findings</div>

- Without culling, an outbreak could cause widespread bird losses across New
  Zealand.

- Acting early through culling could limit the expected national share of birds
  lost to **`r round(cull_national_loss_pct, 1)`%**, compared with
  **`r round(mean_loss_share, 1)`% without culling**, under the current settings.

- No two outbreaks unfold in exactly the same way. Response planning should
  therefore allow for a range of possible losses, not rely on one estimate.

- With culling, the expected share of birds lost by production type falls to
  **`r cull_type_loss_comparison`**.

- Without culling, the average chance of a farm being lost is
  **`r round(north$mean_lost_probability, 1)`% in the North Island** and
  **`r round(south$mean_lost_probability, 1)`% in the South Island**. With
  culling, this falls to **`r round(cull_north_probability, 1)`%** and
  **`r round(cull_south_probability, 1)`%**, respectively. The North Island
  remains more exposed because more farms are located there, making a random
  start there more likely, while spread between islands is blocked.

- These findings can help compare response choices and focus preparedness.
  They should not be read as a forecast of a future outbreak.

:::

```{r}
#| label: production-reference-table
#| include: false
capacity_reference_row <- function(label, rows) {
  capacity <- farms$production_yield[rows]
  data.frame(
    `Production type` = label,
    Farms = length(capacity),
    `Total birds` = sum(capacity),
    `Share of all birds` = 100 * sum(capacity) /
      sum(farms$production_yield),
    `Average per farm` = mean(capacity),
    `Median per farm` = median(capacity),
    `Middle half per farm` = paste0(
      thousand_birds(unname(quantile(capacity, .25))), " to ",
      thousand_birds(unname(quantile(capacity, .75)))),
    `Minimum to maximum per farm` = paste0(
      thousand_birds(min(capacity)), " to ",
      thousand_birds(max(capacity))),
    check.names = FALSE)
}
capacity_reference <- do.call(rbind, c(
  lapply(sort(unique(farms$production_type)), function(type) {
    capacity_reference_row(type, farms$production_type == type)
  }),
  list(capacity_reference_row("Grand total", rep(TRUE, nrow(farms))))))
capacity_reference$`Total birds` <- format(
  round(capacity_reference$`Total birds` / 1000, 1), nsmall = 1, big.mark = ",")
capacity_reference$`Share of all birds` <- paste0(
  round(capacity_reference$`Share of all birds`, 1), "%")
capacity_reference$`Average per farm` <- thousand_birds(
  capacity_reference$`Average per farm`)
capacity_reference$`Median per farm` <- thousand_birds(
  capacity_reference$`Median per farm`)
```

## Culling reduces bird losses nationally and across production types

@tbl-expected-population-loss compares bird losses under two management
responses. **No culling is the reference** and **culling is the response
scenario**. With culling, the national share of birds lost falls from
**`r round(mean_loss_share, 1)`% to `r round(cull_national_loss_pct, 1)`%**.
The share also falls for both Eggs and Meat. The Total row shows the national
result, while the production-type rows show where those losses occur.

```{r}
#| label: tbl-expected-population-loss
#| tbl-cap: "Bird losses under no culling and culling"
#| results: asis
make_population_rows <- function(scenario, risk_data, loss_log) {
  rows <- lapply(sort(unique(risk_data$production_type)), function(type) {
    in_type <- risk_data$production_type == type
    total_birds <- sum(risk_data$production_yield[in_type])
    expected_loss <- sum(risk_data$expected_loss_contribution[in_type])
    loss_share <- 100 * loss_log[, type] / total_birds
    data.frame(
      `Management response` = scenario,
      `Production type` = type,
      `Farms in population` = sum(in_type),
      `Birds in population` = total_birds,
      `Expected total birds lost` = expected_loss,
      `Expected share of birds lost` = 100 * expected_loss / total_birds,
      `Share p25` = unname(quantile(loss_share, .25)),
      `Share p75` = unname(quantile(loss_share, .75)),
      check.names = FALSE)
  })
  total_birds <- sum(risk_data$production_yield)
  expected_loss <- sum(risk_data$expected_loss_contribution)
  total_loss_share <- 100 * rowSums(loss_log) / total_birds
  rows[[length(rows) + 1L]] <- data.frame(
    `Management response` = scenario,
    `Production type` = "Total",
    `Farms in population` = nrow(risk_data),
    `Birds in population` = total_birds,
    `Expected total birds lost` = expected_loss,
    `Expected share of birds lost` = 100 * expected_loss / total_birds,
    `Share p25` = unname(quantile(total_loss_share, .25)),
    `Share p75` = unname(quantile(total_loss_share, .75)),
    check.names = FALSE)
  do.call(rbind, rows)
}

population_estimates <- rbind(
  make_population_rows("No culling", farm_risk,
                       results$production_loss_by_type_log),
  make_population_rows("Culling", cull_farm_risk,
                       results$cull_production_loss_by_type_log))
population_estimates$`Birds in population` <- format(
  round(population_estimates$`Birds in population` / 1000, 1),
  nsmall = 1, big.mark = ",")
population_estimates$`Expected total birds lost` <- format(
  round(population_estimates$`Expected total birds lost` / 1000, 1),
  nsmall = 1, big.mark = ",")
population_estimates$`Expected share of birds lost` <- paste0(
  round(population_estimates$`Expected share of birds lost`, 1), "% (",
  round(population_estimates$`Share p25`, 1), "% to ",
  round(population_estimates$`Share p75`, 1), "%)")
population_estimates$`Share p25` <- population_estimates$`Share p75` <- NULL
names(population_estimates)[names(population_estimates) == "Birds in population"] <-
  "Birds in population (thousands)"
names(population_estimates)[names(population_estimates) == "Expected total birds lost"] <-
  "Birds lost (thousands)"
names(population_estimates)[names(population_estimates) == "Expected share of birds lost"] <-
  "Share of birds lost"

escape_html <- function(x) {
  x <- gsub("&", "&amp;", as.character(x), fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  gsub('"', "&quot;", x, fixed = TRUE)
}

headers <- names(population_estimates)
cat('<table class="table table-sm table-striped population-loss-table">\n<thead><tr>')
cat(paste0("<th>", escape_html(headers), "</th>", collapse = ""))
cat("</tr></thead>\n<tbody>\n")
for (response in unique(population_estimates$`Management response`)) {
  group_rows <- population_estimates[
    population_estimates$`Management response` == response, , drop = FALSE]
  for (i in seq_len(nrow(group_rows))) {
    cat("<tr>")
    if (i == 1L) {
      cat(sprintf('<td rowspan="%d">%s</td>', nrow(group_rows),
                  escape_html(response)))
    }
    cat(paste0("<td>", escape_html(unlist(group_rows[i, -1], use.names = FALSE)),
               "</td>", collapse = ""))
    cat("</tr>\n")
  }
}
cat("</tbody></table>\n")
```

```{=html}
<div class="table-note" role="note"><strong>How to read this table:</strong>
Bird counts are shown in thousand birds. The share lost compares birds lost
with all birds in that production type. Percentages in brackets show the middle
half of the modelled outcomes. The calculation is explained in the appendix.</div>
```

Birds removed through culling are included in the loss shown in Table 1. Even
after counting those birds, culling produces a lower total loss in
**`r round(cull$runs_with_lower_total_loss_pct, 1)`% of the modelled
outbreaks**. The result shows the trade-off: some birds are lost through early
culling to prevent a much larger loss across the wider farm population.

The percentages in brackets in Table 1 show the middle half of the modelled
outcomes. They give planners a practical range rather than relying on one
estimate. The appendix shows the full set of repeated outcomes and explains
how the comparison was made.

This does not mean the same saving would occur in a real event. The result says
that a prompt and well-covered response is valuable under the modelled
conditions. Whether that can be achieved depends on detection and operational
capacity; the response assumptions are set out in the appendix.

## Both production types benefit from a culling response

![Birds lost for each production type across the modelled outbreaks, split into No culling and Culling panels on the same scale. Each box describes all birds lost in that category, not an average farm.](assets/spatial_cascade_loss_by_production_type.svg){#fig-loss-by-type fig-alt="Birds lost by production type with no culling and with culling"}

Both production types follow the same broad national pattern: losses are high
without a response and lower with culling. A category with more birds exposed
can carry a larger absolute loss. Production type is used here to describe
where losses fall; it does not change how spread occurs in the model. Table 1
provides the detailed production-type percentages and ranges.

## Shelter and culling affect how many birds remain unaffected

![Culling leaves more birds unaffected at sheltered and unsheltered farms.](assets/spatial_cascade_survival_comparison.svg){#fig-shelter-comparison fig-alt="Share of birds at sheltered and unsheltered farms remaining unaffected by production type and national total, with no culling and with culling"}

```{=html}
<div class="figure-note" role="note"><strong>How to read this figure:</strong>
Higher bars mean that a larger share of birds remains unaffected. Eggs and Meat
show the results for those production types; Total shows them together. The
left panel has no culling and the right panel includes culling. Birds at culled
farms are counted as lost. <strong>What it means:</strong> Culling leaves a much
larger share of birds unaffected. The difference between sheltered and
unsheltered birds is not the same for Eggs and Meat, so the chart should not be
read as evidence that shelter alone caused the difference.</div>
```

Without culling, the share of birds lost is
**`r round(100 - summary$sheltered_bird_survival_pct, 1)`% at sheltered farms**
and **`r round(100 - summary$unsheltered_bird_survival_pct, 1)`% at unsheltered
farms**. With culling, the shares lost fall to
**`r round(100 - cull_average_bird_survival[[config$scenario_metadata$sheltered_label]], 1)`%**
and **`r round(100 - cull_average_bird_survival[[config$scenario_metadata$unsheltered_label]], 1)`%**,
respectively. These are shares of birds, not shares of farms.

Shelter lowers the chance that a farm is reached in the model. However, the
bars also reflect where sheltered farms are located, how many birds they hold
and where outbreaks begin. This helps explain why the shelter pattern differs
between Eggs and Meat. The appendix explains how these shares are calculated.

## Some farming areas are exposed more often than others

![Farm exposure across New Zealand without culling and with culling. Larger bubbles represent more birds. Colour shows how often all birds at each farm were counted as lost across the modelled outbreaks.](assets/spatial_cascade_farm_risk_map.svg){#fig-farm-risk fig-alt="New Zealand farm bird-loss exposure without culling and with culling"}

@fig-farm-risk shows the spatial distribution of the simulation results across
New Zealand. Each bubble is a farm at its fixed location. Larger bubbles hold
more birds, while darker colours identify farms that are lost more often across
the modelled outbreaks. The left map shows outcomes without culling and the
right map shows outcomes with culling. Both maps use the same colour scale, so
their patterns can be compared directly.

On the culling map, a farm is counted as lost if it is deliberately culled or
if it reaches the lost state through the outbreak. A dark bubble on the right
does not therefore always represent uncontrolled spread. It may show an early
cull that helped protect farms elsewhere.

A value of 40% means that the farm was lost in 40 out of every 100 modelled
outbreaks under that response. In those outbreaks, all birds at the farm are
counted as lost. The percentage shows repeated exposure under the selected
settings; it is not a forecast that a real outbreak will reach that farm.

Across all farms, the average chance of being lost falls from
**`r round(mean(farm_risk$probability_lost_pct), 1)`% without control** to
**`r round(mean(cull_farm_risk$probability_lost_pct), 1)`% with culling**.
On average, farms are directly culled in
**`r round(mean(cull_farm_risk$probability_culled_pct), 1)`% of modelled outbreaks**.
The culling map therefore combines the farms deliberately removed by the
response with the smaller outbreak chain that continues despite that action.

### More outbreaks begin in the North Island

In the current results, the average chance of reaching the lost state is
**`r round(north$mean_lost_probability, 1)`% for North Island farms** and
**`r round(south$mean_lost_probability, 1)`% for South Island farms**. The main
reason is where the modelled outbreaks begin.

There are `r north$farms` North Island farms and `r south$farms` South Island
farms in the fixed population. With `r initial_starts` starting farms chosen at
random, at least one North Island farm starts the event in about
**`r round(north$start_probability, 1)`% of runs**. The South Island receives a
starting farm in about **`r round(south$start_probability, 1)`% of runs**.

The cross-island setting is currently 0, so an event cannot move from one
island to the other. If a run has no South Island starting farm, every South
Island farm stays unaffected in that run. The larger North Island population
therefore gives the North Island many more opportunities to be involved.

This should not be read as proof that North Island farms are naturally more
vulnerable. It is mainly a result of the number and location of farms, random
starting points, and the blocked island crossing. The shelter shares are also
fairly similar—`r round(north$sheltered_share, 1)`% in the North Island and
`r round(south$sheltered_share, 1)`% in the South Island—so shelter mix is not
the main explanation here. These figures will update automatically when the
full farm population is loaded and the analysis is rerun.

# Appendix: Method and use

## Overview of the simulation

The simulation follows a harmful event through a fixed population of farms.
Each farm keeps its real location and recorded attributes, including its bird
count, production type and shelter category. Each modelled outbreak begins at
`r config$simulation_controls$initial_phase_c_entities` farms selected at
random and can reach other farms, with nearby farms generally more exposed
than distant farms.

A single event can develop in many ways, so the analysis considers many
plausible scenarios rather than relying on one chain. Those scenarios show the
central result, the range of outcomes and the farms that are repeatedly
exposed. Each event is also considered with and without culling. This creates
a fair management comparison because both versions begin from the same places
and face the same underlying chance conditions.

The scenarios are not forecasts of what will happen. They are a structured
way to understand possible scale, location and uncertainty, and to test how
much a management response could change the outcome.

## Farm-level descriptive context

This is the only table in the report that describes a typical **individual
farm**. It is supporting context, not the national result. “Average per farm”
means the total birds in a category divided by the number of farms in that
category.

```{r}
#| label: farm-descriptive-table
#| tbl-cap: "Bird-count distribution across individual farms"
farm_context <- capacity_reference[, c("Production type", "Farms",
  "Average per farm", "Median per farm", "Middle half per farm",
  "Minimum to maximum per farm")]
names(farm_context)[3:6] <- c(
  "Average per farm (thousand birds)",
  "Median per farm (thousand birds)",
  "Middle half per farm (thousand birds)",
  "Minimum to maximum per farm (thousand birds)")
knitr::kable(farm_context, row.names = FALSE)
```

## Repeated scenarios and the paired comparison

One run gives only one possible chain of events. This analysis repeats the
whole scenario **`r format(runs, big.mark = ",")` times** over the same fixed
farm population. Starting farms and chance outcomes can differ between runs.
Looking across all runs shows the usual result as well as less common severe
outcomes.

More runs make the reported pattern steadier; they do not make the assumptions
more accurate. Results still depend on the farm data and selected settings.

Each scenario begins with every farm unaffected and
`r config$simulation_controls$initial_phase_c_entities` farms starting the
event. The model follows the next `r days` days. A farm may be reached,
incubate, become active and then have all its birds counted as lost.

Every starting pattern is run twice. The no-culling and culling versions use
the same farms and prepared random values. This pairing keeps the comparison
focused on the management response rather than unrelated random differences.

![Repeated outbreaks show why national losses are reported as a range.](assets/spatial_cascade_loss_distribution.svg){#fig-loss-distribution fig-alt="Percentage of birds lost and remaining in repeated modelled outbreaks with no culling and with culling"}

```{=html}
<div class="figure-note" role="note"><strong>How to read this figure:</strong>
Each thin bar is one complete modelled outbreak and represents 100% of birds in
the national farm population. Orange shows birds lost and grey shows birds
remaining unaffected. Bars are ordered from lower to higher loss within each
panel; this is not a timeline. <strong>What it means:</strong> Outcomes vary
even when the farm population and model settings stay the same. The larger
grey area with culling shows that the response reduces losses across most of
the repeated outbreaks.</div>
```

Without culling, nine out of ten modelled outbreaks finished between
**`r thousand_birds(summary$loss_p05)` and
`r thousand_birds(summary$loss_p95)` thousand birds lost**.
This range is why the report presents both a central estimate and a spread of
possible outcomes.

Under the current culling settings, **`r round(cull$coverage_pct)`% of active
farms are identified for response** and are culled
**`r cull$response_delay_days` day after becoming active**. There is no separate
daily cap. All birds at a
culled farm count as a direct loss, while the farm stops contributing to later
spread.

## How the bird shares are calculated

The main loss and survival results are based on bird numbers. For each modelled
outbreak, the analysis adds the birds at farms that are lost and divides that
number by all birds in the same group. The share remaining unaffected is 100%
minus the share lost. Birds at culled farms are included in the loss.

Groups are kept separate. For example, the Eggs–Sheltered result uses birds at
sheltered egg farms only. The Total–Sheltered result combines Eggs and Meat but
still uses sheltered farms only. The same calculation is repeated for
unsheltered farms and for each management response.

The final percentage combines the results from the repeated outbreaks. It is
not an average farm result: each farm contributes according to the number of
birds it holds. Farm counts are reported separately where they are useful for
response planning.

## Timing and distance

![The left panel follows one farm after it is reached and shows when it can spread, when it would normally move to lost production, and when a covered farm would be culled. The right panel shows how one transmission opportunity becomes less likely as the distance between two farms increases.](assets/spatial_cascade_timing_distance_concept.svg){#fig-timing-distance fig-alt="Two-panel concept figure showing farm state timing and spatial transmission decay"}

The two parts of the model work together. **Timing controls when a farm can
spread. Distance controls how likely it is to reach another farm while it is
active.**

The left panel begins when a farm is reached. It spends
`r config$model_parameters$incubation_period_days` days incubating and cannot
spread during that period. It then becomes active for up to
`r config$model_parameters$active_period_days` days. Without control, it moves
to lost production at the end of that active period. If the farm is covered by
the culling policy, it is culled
`r config$cull_response$response_delay_days` day after becoming active and can
no longer spread. Coverage still matters: only
`r round(100 * config$cull_response$coverage)`% of active farms are selected
for that response under the current settings.

The right panel begins with one active farm and one unaffected farm. At short
distances, the model gives that transmission opportunity a higher chance. The
chance falls smoothly as distance increases. Shelter lowers the whole curve
for a sheltered farm. The different-island curve also includes the island
setting; with the current value of
`r config$model_parameters$cross_barrier_transmission_multiplier`, it stays at
zero because cross-island spread is blocked.

This curve describes one opportunity from one active source. When several
active farms can reach the same farm on the same day, the model combines those
opportunities, so the farm's overall chance of being reached can be higher than
one line alone suggests.

## Movement between islands

The farm data identify which island each farm belongs to. Farms on the same
island have no extra barrier. For farms on different islands, the model applies
the cross-island setting after allowing for distance and shelter.

The current value is
**`r config$model_parameters$cross_barrier_transmission_multiplier`**. A value
of 0 blocks spread between islands. A value such as 0.05 leaves a small
cross-island pathway, while 1 removes the extra island penalty but still allows
distance to reduce spread. A value above 0 should only be used when there is a
credible route, such as ferry movements, shared equipment or supplier links.

The grey New Zealand shape and optional regional lines help readers locate the
farms. They do not change the calculation.

## Appropriate use and limitations

Use the results to compare plausible management settings, understand the
possible national scale of loss and identify places that are repeatedly
exposed. The expected whole-population loss is the clearest headline for
communication; farm averages provide context but answer a different question.

The model simplifies real pathways. Straight-line distance cannot fully
represent movements of birds, staff, vehicles or shared equipment, nor wind,
cleaning practice and operational constraints. Results depend on the quality
of the farm table and the selected assumptions. They should support planning
and discussion, not replace field evidence or operational judgement.

## Running the analysis again

Start with the complete farm table in `data/farms.csv`. If column names change,
update the mappings in `config.json`. Review the starting-farm, timing,
distance, shelter, island, culling and run settings, then run
`source("spatial_cascade_monte_carlo.R")`. Refreshed figures and tables are
written to `assets/`. Run `quarto render spatial_cascade_report.qmd` to rebuild
this report.

The analysis also writes farm-level risk files to `assets/`. These record how
often each farm begins an outbreak, becomes affected, is lost or is culled,
along with its contribution to the expected national loss. Because starting
farms are selected at random, individual farms will not start exactly the same
number of outbreaks. Under the current settings, each farm is selected about
`r round(runs * config$simulation_controls$initial_phase_c_entities / nrow(farms), 1)`
times on average.
