---
title: "Exploring the Potential Spread and Impact of H5N1 in New Zealand Poultry Farms"
subtitle: "Planning for the risk that remains with a culling response"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{=html}
<style>
.table-note,
.figure-note {
  margin-top: -0.55rem;
  margin-bottom: 1.25rem;
  color: #7a8791;
  font-size: 0.82rem;
  line-height: 1.35;
}
.population-loss-table td[rowspan] {
  vertical-align: middle;
  font-weight: 600;
  background: #f5f7f8;
}
.key-findings {
  margin: 1rem 0 1.5rem;
  padding: 1rem 1.2rem 0.35rem;
  background: #fff7d6;
  border: 1px solid #eadb9a;
  border-left: 5px solid #d8bd55;
  border-radius: 0.35rem;
}
.key-findings-title {
  margin-bottom: 0.65rem;
  color: #5f5223;
  font-weight: 700;
}
</style>
```

```{r}
#| label: setup
library(here)
config <- jsonlite::fromJSON("config.json")
results <- readRDS(here("assets", "spatial_cascade_results.rds"))
summary <- read.csv(here("assets", "spatial_cascade_summary.csv"))
farms <- results$farms
runs <- results$monte_carlo_runs
days <- config$model_parameters$simulation_days
capacity_unit <- config$production_metrics$production_capacity_unit
if (is.null(capacity_unit) || !length(capacity_unit)) capacity_unit <- "birds"
thousand_birds <- function(x, digits = 1) {
  format(round(x / 1000, digits), nsmall = digits, big.mark = ",", trim = TRUE)
}
mean_loss_share <- 100 * summary$mean_production_loss /
  sum(farms$production_yield)
cull <- results$cull_comparison_summary
farm_risk <- results$farm_risk_summary
cull_farm_risk <- results$cull_farm_risk_summary
cull_national_loss_pct <- 100 * cull$mean_cull_total_loss /
  sum(farms$production_yield)
cull_national_loss_runs <- 100 *
  rowSums(results$cull_production_loss_by_type_log) /
  sum(farms$production_yield)
cull_national_loss_p25 <- unname(quantile(cull_national_loss_runs, 0.25))
cull_national_loss_p75 <- unname(quantile(cull_national_loss_runs, 0.75))
cull_average_bird_survival <- results$cull_average_bird_survival
north_name <- grep("^North", unique(farms$barrier_group), value = TRUE)[1]
south_name <- grep("^South", unique(farms$barrier_group), value = TRUE)[1]
initial_starts <- config$simulation_controls$initial_phase_c_entities
island_story <- function(group_name) {
  rows <- farms$barrier_group == group_name
  group_size <- sum(rows)
  start_probability <- 100 * (1 -
    choose(nrow(farms) - group_size, initial_starts) /
      choose(nrow(farms), initial_starts))
  group_risk <- mean(farm_risk$probability_lost_pct[
    farm_risk$barrier_group == group_name])
  sheltered_share <- 100 * mean(farms$is_sheltered[rows])
  data.frame(group = group_name, farms = group_size,
    start_probability = start_probability, mean_lost_probability = group_risk,
    sheltered_share = sheltered_share)
}
island_comparison <- rbind(island_story(north_name), island_story(south_name))
north <- island_comparison[island_comparison$group == north_name, ]
south <- island_comparison[island_comparison$group == south_name, ]
cull_north_probability <- mean(cull_farm_risk$probability_lost_pct[
  cull_farm_risk$barrier_group == north_name])
cull_south_probability <- mean(cull_farm_risk$probability_lost_pct[
  cull_farm_risk$barrier_group == south_name])
production_types <- sort(unique(farms$production_type))
cull_type_loss_pct <- vapply(production_types, function(type) {
  type_birds <- sum(farms$production_yield[farms$production_type == type])
  expected_loss <- sum(cull_farm_risk$expected_loss_contribution[
    cull_farm_risk$production_type == type])
  100 * expected_loss / type_birds
}, numeric(1))
cull_type_loss_order <- order(cull_type_loss_pct, decreasing = TRUE)
cull_type_loss_comparison <- paste(
  paste0(tolower(production_types[cull_type_loss_order]), ": ",
         round(cull_type_loss_pct[cull_type_loss_order], 1), "%"),
  collapse = "; ")
```

::: {.key-findings}

<div class="key-findings-title">Key findings</div>

- The expected national share of birds lost is
  **`r round(cull_national_loss_pct, 1)`%**.

- Bird loss is similar across production types:
  **`r cull_type_loss_comparison`**.

- The chance of a farm being affected by an outbreak is
  `r round(cull_north_probability, 1)`% in the North Island and
  `r round(cull_south_probability, 1)`% in the South Island.
  The North Island remains more exposed because more farms are located there,
  making a random start there more likely, while spread between islands is
  blocked.

- These findings can help compare response choices and focus preparedness.
  They should not be read as a forecast of a future outbreak.

:::

```{r}
#| label: production-reference-table
#| include: false
capacity_reference_row <- function(label, rows) {
  capacity <- farms$production_yield[rows]
  data.frame(
    `Production type` = label,
    Farms = length(capacity),
    `Total birds` = sum(capacity),
    `Share of all birds` = 100 * sum(capacity) /
      sum(farms$production_yield),
    `Average per farm` = mean(capacity),
    `Median per farm` = median(capacity),
    `Middle half per farm` = paste0(
      thousand_birds(unname(quantile(capacity, .25))), " to ",
      thousand_birds(unname(quantile(capacity, .75)))),
    `Minimum to maximum per farm` = paste0(
      thousand_birds(min(capacity)), " to ",
      thousand_birds(max(capacity))),
    check.names = FALSE)
}
capacity_reference <- do.call(rbind, c(
  lapply(sort(unique(farms$production_type)), function(type) {
    capacity_reference_row(type, farms$production_type == type)
  }),
  list(capacity_reference_row("Grand total", rep(TRUE, nrow(farms))))))
capacity_reference$`Total birds` <- format(
  round(capacity_reference$`Total birds` / 1000, 1), nsmall = 1, big.mark = ",")
capacity_reference$`Share of all birds` <- paste0(
  round(capacity_reference$`Share of all birds`, 1), "%")
capacity_reference$`Average per farm` <- thousand_birds(
  capacity_reference$`Average per farm`)
capacity_reference$`Median per farm` <- thousand_birds(
  capacity_reference$`Median per farm`)
```

## Around one in five birds remains at risk nationally

The culling response is the main planning case because an affected farm would
realistically trigger a response. No culling is included only as a benchmark
to show how much the response changes the outcome. Details of the simulation
and comparison are in the appendix.

The main planning scenario leaves an expected national loss of
**`r round(cull_national_loss_pct, 1)`% of birds**.
@tbl-expected-population-loss places this result in context by showing the
national total, the losses across eggs and meat, and the no-response benchmark.

```{r}
#| label: tbl-expected-population-loss
#| tbl-cap: "Around one in five birds remains at risk nationally, with similar losses across eggs and meat"
#| results: asis
make_population_rows <- function(scenario, risk_data, loss_log) {
  rows <- lapply(sort(unique(risk_data$production_type)), function(type) {
    in_type <- risk_data$production_type == type
    total_birds <- sum(risk_data$production_yield[in_type])
    expected_loss <- sum(risk_data$expected_loss_contribution[in_type])
    loss_share <- 100 * loss_log[, type] / total_birds
    data.frame(
      `Management response` = scenario,
      `Production type` = type,
      `Farms in population` = sum(in_type),
      `Birds in population` = total_birds,
      `Expected total birds lost` = expected_loss,
      `Expected share of birds lost` = 100 * expected_loss / total_birds,
      `Share p25` = unname(quantile(loss_share, .25)),
      `Share p75` = unname(quantile(loss_share, .75)),
      check.names = FALSE)
  })
  total_birds <- sum(risk_data$production_yield)
  expected_loss <- sum(risk_data$expected_loss_contribution)
  total_loss_share <- 100 * rowSums(loss_log) / total_birds
  rows[[length(rows) + 1L]] <- data.frame(
    `Management response` = scenario,
    `Production type` = "Total",
    `Farms in population` = nrow(risk_data),
    `Birds in population` = total_birds,
    `Expected total birds lost` = expected_loss,
    `Expected share of birds lost` = 100 * expected_loss / total_birds,
    `Share p25` = unname(quantile(total_loss_share, .25)),
    `Share p75` = unname(quantile(total_loss_share, .75)),
    check.names = FALSE)
  do.call(rbind, rows)
}
population_estimates <- rbind(
  make_population_rows("Culling", cull_farm_risk,
                       results$cull_production_loss_by_type_log),
  make_population_rows("No culling", farm_risk,
                       results$production_loss_by_type_log))
population_estimates$`Birds in population` <- format(
  round(population_estimates$`Birds in population` / 1000, 1),
  nsmall = 1, big.mark = ",")
population_estimates$`Expected total birds lost` <- format(
  round(population_estimates$`Expected total birds lost` / 1000, 1),
  nsmall = 1, big.mark = ",")
population_estimates$`Expected share of birds lost` <- paste0(
  round(population_estimates$`Expected share of birds lost`, 1), "% (",
  round(population_estimates$`Share p25`, 1), "% to ",
  round(population_estimates$`Share p75`, 1), "%)")
population_estimates$`Share p25` <- population_estimates$`Share p75` <- NULL
names(population_estimates)[names(population_estimates) == "Birds in population"] <-
  "Birds in population (thousands)"
names(population_estimates)[names(population_estimates) == "Expected total birds lost"] <-
  "Expected birds lost (thousands)"
escape_html <- function(x) {
  x <- gsub("&", "&amp;", as.character(x), fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  gsub('"', "&quot;", x, fixed = TRUE)
}
headers <- names(population_estimates)
cat('<table class="table table-sm table-striped population-loss-table">\n<thead><tr>')
cat(paste0("<th>", escape_html(headers), "</th>", collapse = ""))
cat("</tr></thead>\n<tbody>\n")
for (response in unique(population_estimates$`Management response`)) {
  group_rows <- population_estimates[
    population_estimates$`Management response` == response, , drop = FALSE]
  for (i in seq_len(nrow(group_rows))) {
    cat("<tr>")
    if (i == 1L) {
      cat(sprintf('<td rowspan="%d">%s</td>', nrow(group_rows),
                  escape_html(response)))
    }
    cat(paste0("<td>", escape_html(unlist(group_rows[i, -1], use.names = FALSE)),
               "</td>", collapse = ""))
    cat("</tr>\n")
  }
}
cat("</tbody></table>\n")
```

```{=html}
<div class="table-note" role="note">Bird counts are shown in thousand birds. The expected share is the expected national loss divided by the birds in that production type. Percentages in brackets show the 25th–75th percentile
range across complete modelled outbreaks with the assumptions held fixed. When
a farm is lost, all birds at that farm are counted as lost in that outbreak.</div>
```

Birds removed at identified farms are included in the national loss. Even after
counting these birds, the response produces a lower total loss in
**`r round(cull$runs_with_lower_total_loss_pct, 1)`%** of outbreaks. The number
of affected farms falls from **`r round(cull$mean_baseline_farms_affected, 1)`**
in the benchmark to **`r round(cull$mean_cull_farms_affected, 1)`** in the main
scenario.

The production-type results are close—**`r cull_type_loss_comparison`**—so the
remaining risk is not confined to either eggs or meat. The number of birds lost
can nevertheless differ because the two production types do not hold the same
number of birds.

The result also varies between outbreaks. Across the middle half of outcomes,
the national share lost ranges from **`r round(cull_national_loss_p25, 1)`% to
`r round(cull_national_loss_p75, 1)`%**. Preparedness therefore needs to account
for outcomes above and below the central estimate.

## Shelter is not associated with the same outcome for eggs and meat

![The share of birds remaining unaffected differs by shelter status and production type.](assets/spatial_cascade_survival_comparison.svg){#fig-shelter-comparison fig-alt="Share of birds at sheltered and unsheltered farms remaining unaffected by production type, with the main response scenario first and no response as a benchmark"}

```{=html}
<div class="figure-note" role="note">
Higher bars mean that a larger share of birds remains unaffected. The main
response scenario is shown first and the no-response benchmark second. Birds
at culled farms are counted as lost. Results are weighted by bird numbers, not
by the number of farms.</div>
```

@fig-shelter-comparison shows that the relationship between shelter and birds
remaining unaffected is not consistent across eggs and meat. Nationally,
**`r round(cull_average_bird_survival[[config$scenario_metadata$sheltered_label]], 1)`%**
of birds at sheltered farms and
**`r round(cull_average_bird_survival[[config$scenario_metadata$unsheltered_label]], 1)`%**
of birds at unsheltered farms remain unaffected.

The shelter pattern differs between eggs and meat. This indicates that shelter
status alone does not explain the result: farm location and the number of birds
held at those farms also shape the observed pattern.

## Farm exposure remains concentrated in the North Island

![Farm exposure remains concentrated in the North Island.](assets/spatial_cascade_farm_risk_map.svg){#fig-farm-risk fig-alt="New Zealand farm exposure with the main response scenario first and no response as a benchmark"}

```{=html}
<div class="figure-note" role="note">
The main response scenario is shown first and the no-response benchmark second.
Each bubble is one farm, larger bubbles hold more birds, and darker colours
mean the farm is lost in a larger share of outbreaks. A culled farm is counted
as lost.</div>
```

@fig-farm-risk shows that exposure remains uneven. The strongest concentration
is in the North Island, although exposure is substantially lower than in the
benchmark.

The national average chance of a farm being lost is
**`r round(mean(cull_farm_risk$probability_lost_pct), 1)`%**.

Farms are directly culled in
**`r round(mean(cull_farm_risk$probability_culled_pct), 1)`%** of outbreaks on
average.

### North Island farms carry more of the remaining exposure

The average chance of a farm being lost is
**`r round(cull_north_probability, 1)`% in the North Island** and
**`r round(cull_south_probability, 1)`% in the South Island**. The difference
is associated mainly with the larger concentration of farms and starting
events in the North Island. Shelter coverage is similar across the islands—
**`r round(north$sheltered_share, 1)`%** in the North Island and
**`r round(south$sheltered_share, 1)`%** in the South Island—so it does not
explain the geographic gap.

# Appendix: Method and use

## Overview of the simulation

The simulation follows a H5N1 outbreak event through a fixed population of farms.

Each farm keeps its real location and recorded attributes, including its bird
count, production type and shelter category. Each modelled outbreak begins at
`r config$simulation_controls$initial_phase_c_entities` farms selected at
random and can reach other farms, with nearby farms generally more exposed
than distant farms.

A single event can develop in many ways, so the analysis considers a range of
possible outbreaks rather than relying on one chain.

### Repeated scenarios and the paired comparison

One run gives only one possible chain of events. This analysis repeats the
whole scenario **`r format(runs, big.mark = ",")` times** over the same fixed
farm population. Starting farms and chance outcomes can differ between runs.

Looking across all runs shows the usual result as well as less common severe
outcomes. More runs make this pattern steadier; they do not make the model
assumptions more accurate.

Each scenario begins with every farm unaffected and
`r config$simulation_controls$initial_phase_c_entities` farms starting the
event. The model follows the next `r days` days. A farm may be reached,
incubate, become active and then have all its birds counted as lost.

Every starting pattern is run twice. The no-culling and culling versions use
the same farms and prepared random values. This pairing keeps the comparison
focused on the management response rather than unrelated random differences.

![Repeated outbreaks show why national losses are reported as a range.](assets/spatial_cascade_loss_distribution.svg){#fig-loss-distribution fig-alt="Percentage of birds lost and remaining in repeated modelled outbreaks with no culling and with culling"}

```{=html}
<div class="figure-note" role="note">
Each thin bar is one complete modelled outbreak and represents 100% of birds in
the national farm population. Orange is calculated by adding the birds at lost
or culled farms and dividing by all birds; green is the share remaining. This
weights farms by their bird numbers rather than treating every farm equally.
The same calculation is used within each production type or shelter group,
using only the birds in that group. Bars are ordered from lower to higher loss
within each panel; this is not a timeline.</div>
```

Outcomes vary even though the farm population and settings remain fixed. In the
main response scenario, the middle half of outbreaks lose between
**`r round(cull_national_loss_p25, 1)`% and
`r round(cull_national_loss_p75, 1)`%** of birds. The larger green area in the
first panel shows that the response leaves more birds unaffected across most
of the repeated outbreaks.

### Timing and distance

![The left panel follows one farm after it is reached and shows when it can spread, when it would normally move to lost production, and when a covered farm would be culled. The right panel shows how one transmission opportunity becomes less likely as the distance between two farms increases.](assets/spatial_cascade_timing_distance_concept.svg){#fig-timing-distance fig-alt="Two-panel concept figure showing farm state timing and spatial transmission decay"}

The two parts of the model work together. Timing controls when a farm can
spread. Distance controls how likely it is to reach another farm while it is
active.

The left panel begins when a farm is reached. It spends
`r config$model_parameters$incubation_period_days` days incubating and cannot
spread during that period. It then becomes active for up to
`r config$model_parameters$active_period_days` days. Without control, it moves
to lost production at the end of that active period. If the farm is covered by
the culling policy, it is culled
`r config$cull_response$response_delay_days` day after becoming active and can
no longer spread. Coverage still matters: only
`r round(100 * config$cull_response$coverage)`% of active farms are selected
for that response under the current settings.

The right panel begins with one active farm and one unaffected farm. At short
distances, the model gives that transmission opportunity a higher chance. The
chance falls smoothly as distance increases. Shelter lowers the whole curve,
while the different-island curve also reflects the island setting described
in the next section.

This curve describes one opportunity from one active source. When several
active farms can reach the same farm on the same day, the model combines those
opportunities, so the farm's overall chance of being reached can be higher than
one line alone suggests.

### Movement between islands

The farm data identify which island each farm belongs to. Farms on the same
island have no extra barrier. For farms on different islands, the model applies
the cross-island setting after allowing for distance and shelter.

The current value is
**`r config$model_parameters$cross_barrier_transmission_multiplier`**. A value
of 0 blocks spread between islands. A value such as 0.05 leaves a small
cross-island pathway, while 1 removes the extra island penalty but still allows
distance to reduce spread. A value above 0 should only be used when there is a
credible route, such as ferry movements, shared equipment or supplier links.

## Farm-level descriptive context

@tbl-farm-descriptive-table describes a typical individual
farm in the dataset.

```{r}
#| label: tbl-farm-descriptive-table
#| tbl-cap: "Bird-count distribution across individual farms"
farm_context <- capacity_reference[, c("Production type", "Farms",
  "Average per farm", "Median per farm", "Middle half per farm",
  "Minimum to maximum per farm")]
names(farm_context)[3:6] <- c(
  "Average per farm (thousand birds)",
  "Median per farm (thousand birds)",
  "Middle half per farm (thousand birds)",
  "Minimum to maximum per farm (thousand birds)")
knitr::kable(farm_context, row.names = FALSE)
```

## Appropriate use and limitations

Use the results to compare plausible management settings, understand the
possible national scale of loss and identify places that are repeatedly
exposed. The expected whole-population loss is the clearest headline for
communication; farm averages provide context but answer a different question.

The model simplifies real pathways. Straight-line distance cannot fully
represent movements of birds, staff, vehicles or shared equipment, nor wind,
cleaning practice and operational constraints. Results depend on the quality
of the farm table and the selected assumptions. They should support planning
and discussion, not replace field evidence or operational judgement.

## Running the analysis again

Start with the complete farm table in `data/map_farm_data.csv`. If column names change,
update the mappings in `config.json`. Review the starting-farm, timing,
distance, shelter, island, culling and run settings, then run
`source("spatial_cascade_monte_carlo.R")`. Refreshed figures and tables are
written to `assets/`. Run `quarto render spatial_cascade_report.qmd` to rebuild
this report.

The analysis also writes farm-level risk files to `assets/`. These record how
often each farm begins an outbreak, becomes affected, is lost or is culled,
along with its contribution to the expected national loss. Because starting
farms are selected at random, individual farms will not start exactly the same
number of outbreaks. Under the current settings, each farm is selected about
`r round(runs * config$simulation_controls$initial_phase_c_entities / nrow(farms), 1)`
times on average.
