---
title: "How Production Loss Could Spread Across New Zealand Farms"
subtitle: "A practical account of the scenario and its results"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
config <- jsonlite::fromJSON("config.json")
results <- readRDS("assets/spatial_cascade_results.rds")
summary <- read.csv("assets/spatial_cascade_summary.csv")
types <- read.csv("assets/spatial_cascade_production_type_summary.csv")
farms <- results$farms
runs <- results$monte_carlo_runs
days <- config$model_parameters$simulation_days
capacity_unit <- config$production_metrics$production_capacity_unit
if (is.null(capacity_unit) || !length(capacity_unit)) capacity_unit <- "birds"
mean_loss_share <- 100 * summary$mean_production_loss /
  sum(farms$production_yield)
protection_difference <- summary$protection_benefit_percentage_points
farm_risk <- results$farm_risk_summary
north_name <- grep("^North", unique(farms$barrier_group), value = TRUE)[1]
south_name <- grep("^South", unique(farms$barrier_group), value = TRUE)[1]
initial_starts <- config$simulation_controls$initial_phase_c_entities

island_story <- function(group_name) {
  rows <- farms$barrier_group == group_name
  group_size <- sum(rows)
  start_probability <- 100 * (1 -
    choose(nrow(farms) - group_size, initial_starts) /
      choose(nrow(farms), initial_starts))
  group_risk <- mean(farm_risk$probability_lost_pct[
    farm_risk$barrier_group == group_name])
  sheltered_share <- 100 * mean(farms$is_sheltered[rows])
  data.frame(group = group_name, farms = group_size,
    start_probability = start_probability, mean_lost_probability = group_risk,
    sheltered_share = sheltered_share)
}

island_comparison <- rbind(island_story(north_name), island_story(south_name))
north <- island_comparison[island_comparison$group == north_name, ]
south <- island_comparison[island_comparison$group == south_name, ]
```

## The story this model tells

This analysis asks a practical question: if a harmful event begins on a small
number of farms, how far could it spread and how much production could be lost?
It follows the real farm locations and attributes supplied in
`r config$input_data$file`.

The current population contains **`r format(nrow(farms), big.mark = ",")` farms**.
Each dot on the map is one of those farms. Its location, region,
production type, shelter category and production capacity stay the same in
every scenario. Production type is used to describe the results. Shelter also
changes the modelled chance of a farm being reached.

At the start of a scenario, every farm is unaffected. The event begins at
`r config$simulation_controls$initial_phase_c_entities` selected farms and the
model follows the next `r days` days. A farm can be reached, spend time in an
incubation period, become able to affect other farms, and finally have its
production counted as lost. Nearby farms are more likely to be reached than
distant farms, although the outcome still includes chance.

## Why the scenario is repeated

One run gives only one possible chain of events. This analysis repeats the
whole scenario **`r format(runs, big.mark = ",")` times** over the same farm
population. The starting locations and the chance outcomes can differ from one
run to the next. Looking across all runs shows which outcomes are common and
which sit at the more severe end of the range.

More runs make the pattern steadier, but they do not make the assumptions more
realistic. The value of the results still depends on the farm data and the
settings used.

## What happened across the whole farm population

![Total production-capacity loss across `r format(runs, big.mark = ",")` complete `r days`-day simulations of all `r format(nrow(farms), big.mark = ",")` farms in the fixed input population. The horizontal axis is the total `r capacity_unit` lost across the whole population in one run; the vertical axis is the number of runs with a similar loss. The dashed line is the average across runs.](assets/spatial_cascade_loss_distribution.svg){#fig-loss-distribution fig-alt="Histogram of total production capacity lost across complete Monte Carlo runs over the full fixed farm population"}

The easiest way to read the chart is to choose one bar and say: **this many
complete scenarios finished with a total population loss around this level**.
The value combines the loss from every farm. It is not an average loss for one
farm, and it is not the chance of a real event beginning.

Across the repeated scenarios, average total loss was
**`r format(round(summary$mean_production_loss), big.mark = ",")` `r capacity_unit`**, about
**`r round(mean_loss_share, 1)`%** of the fixed population's production
capacity. Nine out of ten scenarios finished between
**`r format(round(summary$loss_p05), big.mark = ",")` and
`r format(round(summary$loss_p95), big.mark = ",")` `r capacity_unit` lost**.

This is a widespread-loss result under the current settings. The average is
useful for a central planning case, while the upper end of the range shows what
a more severe outcome could look like. Because loss is already high in most
runs, moderate changes to protection may be difficult to distinguish unless
the spread and timing settings are well supported by evidence.

## How the result differs by production type

![Total production capacity lost for each production type across the repeated scenarios. Each box shows the range of whole-population loss for that category, not the loss of an average farm.](assets/spatial_cascade_loss_by_production_type.svg){#fig-loss-by-type fig-alt="Distribution of total production capacity lost by production type"}

Production type does not change how the event spreads in this model. It is used
to show where the production consequences fall. A category with more farms or
more capacity will generally have a larger total loss.

```{r}
#| label: type-table
#| tbl-cap: "What the repeated scenarios mean for each production type"
display <- types[, c("production_type", "farms_in_population",
  "farms_in_population_pct", "average_farms_lost",
  "average_farms_lost_pct", "mean_total_production_loss",
  "total_loss_p25", "total_loss_p75")]
display$farms_in_population_pct <- paste0(round(display$farms_in_population_pct, 1), "%")
display$average_farms_lost <- paste0(round(display$average_farms_lost, 1),
  " (", round(display$average_farms_lost_pct, 1), "%)")
display$average_farms_lost_pct <- NULL
display$mean_total_production_loss <- paste(
  format(round(display$mean_total_production_loss), big.mark = ","), capacity_unit)
display$total_loss_p25 <- paste0(format(round(display$total_loss_p25), big.mark = ","),
  " to ", format(round(display$total_loss_p75), big.mark = ","), " ", capacity_unit)
display$total_loss_p75 <- NULL
names(display) <- c("Production type", "Farms", "Population share",
  "Average farms lost (%)", "Average total capacity lost",
  "Middle half of results")
knitr::kable(display, row.names = FALSE)
```

The “middle half” column leaves out the lowest quarter and highest quarter of
the runs. It gives a simple view of the more usual range without implying that
the more severe results cannot happen.

## What the shelter comparison says

![Average share of sheltered and unsheltered farms that remain unaffected at the end of the scenario, shown by production type.](assets/spatial_cascade_survival_comparison.svg){#fig-shelter-comparison fig-alt="Comparison of sheltered and unsheltered farm survival by production type"}

Sheltered survival averaged **`r round(summary$sheltered_survival_pct, 1)`%**,
compared with **`r round(summary$unsheltered_survival_pct, 1)`%** for
unsheltered farms. The difference was
**`r round(summary$protection_benefit_percentage_points, 1)` percentage
points** under the configured assumptions.

```{r}
#| results: asis
if (protection_difference < 0) {
  cat(paste0(
    "In these results, sheltered farms have a slightly lower raw survival ",
    "rate. This does **not** mean shelter causes harm. The sheltered farms may ",
    "simply be in more exposed places or may have been chosen as starting ",
    "farms more often. The chart describes this farm population; it does not ",
    "prove what shelter caused.\n"
  ))
} else {
  cat(paste0(
    "Sheltered farms have a higher raw survival rate in these results. Some of ",
    "that difference may still come from where those farms are located, so it ",
    "should not be treated as proof of the shelter effect on its own.\n"
  ))
}
```

The cleanest future test would compare the same farms and the same starting
conditions twice: once with shelter protection and once without it. That would
separate the modelled shelter effect from differences in farm location.

## Where the exposure sits

![Farm exposure across New Zealand. Larger bubbles have more production capacity. Colour shows how often each farm reached the lost state across all repeated scenarios.](assets/spatial_cascade_farm_risk_map.svg){#fig-farm-risk fig-alt="Map of fixed New Zealand farms coloured by how often they reach the lost state"}

The map brings both islands into one view. Each bubble is one farm at its fixed
location. Larger bubbles have more production capacity. The colour shows the
share of repeated scenarios in which the farm reached the lost state.

For example, a value of 40% means that the farm reached the lost state in 40
out of every 100 simulated scenarios under the current settings and starting
location method. It does not mean the farm is certain to lose 40% of its own
production, and it is not a forecast that an event will occur. It is a
scenario-based exposure measure: a high value identifies a farm that is often
reached when many possible cascades are tested.

For example, 40% means the farm reached the lost state in 40 out of every 100
modelled scenarios. It does not mean that the farm loses 40% of its own
production, and it is not a forecast that the event will occur. A high value
simply marks a farm that is frequently exposed under the chosen settings and
starting method.

The supporting farm-risk CSV also records how often each farm started a
scenario, how often it was affected, how often it was lost, and its expected
contribution to total loss. In random-start mode, some farms will be selected
as starting points more often than others just by chance. The current setup
selects each farm about
`r round(runs * config$simulation_controls$initial_phase_c_entities / nrow(farms), 1)`
times on average.

### Why the North Island looks more exposed

In the current results, the average chance of reaching the lost state is
**`r round(north$mean_lost_probability, 1)`% for North Island farms** and
**`r round(south$mean_lost_probability, 1)`% for South Island farms**. The main
reason is how the scenarios begin.

There are `r north$farms` North Island farms and `r south$farms` South Island
farms in the fixed population. With `r initial_starts` starting farms chosen at
random, at least one North Island farm starts the event in about
**`r round(north$start_probability, 1)`% of runs**. The South Island receives a
starting farm in about **`r round(south$start_probability, 1)`% of runs**.

The cross-island setting is currently 0, so an event cannot move from one
island to the other. If a run has no South Island starting farm, every South
Island farm stays unaffected in that run. The larger North Island population
therefore gives the North Island many more opportunities to be involved.

This should not be read as proof that North Island farms are naturally more
vulnerable. It is mainly a result of the number and location of farms, random
starting points, and the blocked island crossing. The shelter shares are also
fairly similar—`r round(north$sheltered_share, 1)`% in the North Island and
`r round(south$sheltered_share, 1)`% in the South Island—so shelter mix is not
the main explanation here. These figures will update automatically when the
full farm population is loaded and the analysis is rerun.

## What happens between the islands

The farm data identify which island each farm belongs to. Farms on the same
island have no extra barrier. For farms on different islands, the model applies
the cross-island setting after allowing for distance and shelter.

The current value is
**`r config$model_parameters$cross_barrier_transmission_multiplier`**. A value
of 0 blocks spread between islands. A value such as 0.05 leaves a small
cross-island pathway, while 1 removes the extra island penalty but still allows
distance to reduce spread. A value above 0 should only be used when there is a
credible route, such as ferry movements, shared equipment or supplier links.

The grey New Zealand shape and optional regional lines are there to make the
map easier to read. They do not change the model.

## What to take from these results

Under the current settings, the main message is that loss is widespread in
most scenarios. The range across runs matters because it shows that the final
loss is not fixed, even when the farm population stays the same. The map then
adds the local view by showing which farms are repeatedly exposed.

These results are best used to compare plausible settings and identify where
better evidence is needed. Straight-line distance cannot represent every real
pathway. Animal movements, staff, vehicles, shared equipment, wind, cleaning
practices and response actions may all change the real pattern. The model
should support discussion, not replace operational judgement.

## Running the analysis again

Start with the complete farm table in `data/farms.csv`. If its column names
change, update the mappings in `config.json`. Choose how starting farms are
selected, review the timing, distance, shelter, island and run settings, then
run `source("spatial_cascade_monte_carlo.R")`. The refreshed charts and tables
will be written to `assets/`. Finish by running
`quarto render spatial_cascade_report.qmd` to rebuild this report.
